Fine-scale Data Reveals Neighborhood Inequalities In Effects of Heat on Expressed Mood

### 2021-06-30 doc/temp-sentiment.tex
more tweaks from overleaf

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{Environmental Studies Program, University of Colorado Boulder}
\affil[3]{Department of Geographical Sciences, University of Maryland College Park}
\affil[4]{University of Cape Town}
\affil[5]{School of Business, East China University of Science and Technology}
\affil[6]{University of Colorado Boulder}
\affil[7]{University of British Columbia}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

@@

%Theoretical: frameworks and pathways
Researchers have proposed a number of frameworks that focus on pathways by which temperature can affect mental health \cite{Berry2018Apr, Palinkas2020Apr, BerryETAL2010}. Work on biological mechanisms emphasizes the mental health effects from dehydration and other consequences of maintaining stable body temperature in high heat \cite{Lohmus2018Jul, sadiq_impact_2019}. Additionally, recent studies find nighttime temperatures affect sleep quality, with consequences for mental health \cite{Obradovich2017May, Mullins2019Dec}. Other studies point to linkages between increased temperatures, overall physical health and increases in injuries which can in turn affect mental health \cite{Berry2007, WHO2007}. Finally, exposure to higher temperatures can affect productivity and income \cite{kjellstrom_impact_2016, Burke2015Nov}, leading to second-order impacts on mental health outcomes \cite{Katz1997, CohnETAL2004, BouchamaETAL2007}.

%Empirical: Heat and mental health
Drawing on these frameworks, results from an emerging literature are beginning to document linkages between higher temperatures worsened mental health outcomes. Multiple studies find strong evidence that higher temperatures are associated with increases in suicides in the United States \cite{Burke2018Aug, Mullins2019Dec, Dixon2007May} with other studies finding a similar relationship in locations around the world \cite{Qi2014Dec, Page2007Aug, Likhvar2011Jan}. Higher temperatures are also correlated with increased hospitalizations related to general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} and incidents related specific conditions like bipolar disorder and schizophrenia \cite{Lee2007Jan, Sung2013Feb}. Higher temperatures are also linked to increased mortality among people with pre-existing mental health disorders \cite{Hansen2008Oct}.

@@

While public health data on mental health outcomes is not available at the neighborhood scale, spontaneous, in situ data from the social media platform Twitter can provide an important, inexpensive and timely proxy for mental health at extremely precise spatial and temporal scales. Many studies use the mood expressed in social media posts as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}. For example, studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder (PTSD) in individuals even before their formal diagnosis \cite{Reece2017Oct}. Similar methods applied to posts on Facebook can predict the onset of depression \cite{Eichstaedt2018Oct}. In London, day-to-day changes in mental health indicators derived from the text of tweets were associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}. Across the USA, more positive tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}. Finally, the mood expressed in tweets has been strongly associated with local weather conditions \cite{baylis_weather_2018, hannak_tweetin_2012} and been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}. Thus, posts on Twitter provide an proxy for mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures. To explore how neighborhood conditions affect the relationship between heat and mental heath, we draw on a data set of 243 million geolocated tweets from across the continental United States covering the period 2009-2019, each paired with data on local weather and neighborhood conditions.

@@

\begin{minipage}{.65\textwidth}
\centering
\includegraphics[width=\textwidth]{../res/wbgt-income-race.png}
\captionof{figure}{Relationship between WBGT and mood from text in tweets, moderated by the interaction of race and income.}
\label{fig:race-income}
\end{minipage}\hfill
\begin{minipage}{.35\textwidth}
\centering
\begin{tabular}{| l | l |}
\hline
Terms & AIC \\
\hline
Neither & 277270561 \\
Income & 277182239 \\
Race & 277143579 \\
Both & 277108676 \\
\hline
\end{tabular}
\captionof{table}{Akaike information criterion (AIC) of models with interaction terms for race, income, both race and income, as well as neither race nor income. The model with the best fit by AIC includes both race and income interaction terms.}
\label{tab:aics}
\end{minipage}

Graphing the combined effects of race and income, we find that race has a larger impact on vulnerability than income (see Fig. \ref{fig:race-income}). A fully black neighborhood at any income level will be much more impacted by heat than a poor neighborhood with few black inhabitants. Additionally, neighborhoods that are largely black are also more affected by cold temperatures, although there is no clear effect of income moderating the relationship between cold temperatures and mood. Using AIC as an indicator of how different terms affect model quality (see Tab. \ref{tab:aics}), we find that a model including both terms performs the best, indicating that both race and income matter for the impact of heat on mood. However, including only a race term improves the model more than including only an income term, further supporting the conclusion that race is more important than income for explaining the vulnerability of mood to heat.

@@

\begin{figure}[H]
 \centering
 \includegraphics[width=0.66\linewidth]{../res/comparison_plot.png}
 \caption{Decline in mood as temperature changes from mild to high across different neighborhood types, compared with the impacts of the change from Saturday to Monday, as well as the change associated with Hurricane Sandy.}
 \label{fig:compare}
\end{figure}

@@

One caveat associated with this data is that we were only able to locate the tweets within the census block where the tweet was sent - we did not to infer where the person sending the tweets typically lived or where they had been. Many low-income and Black people commute during the day to work in the service sector in higher-income areas, so these results may in fact under-estimate the impacts of higher temperatures on mental health for poor and minority individuals. Moreover, in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are typically found that census block. For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day. Additionally, neighborhoods of young college students are also typically estimated to have low income levels, even through college students are wealthier than the average American. Again, these issues mean we are likely under-estimating the true effect of heat on mental health for poor and Black people. A final issue with using Twitter data is that, while Twitter is used by more than one in five Americans, Twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}. Nevertheless, we found a large volume of tweets across all neighborhood types.

@@

Where $y$ is the mood of a tweet, $t$ is the wet bulb globe temperature at the hour of the tweet, $p$ is a binary variable indicating whether it rained at the hour of the tweet, $s$ is the income shortwave radiation, or sunshine, in $W/m^2$, at the hour of the tweet, $\Phi$ is the spatio-temporal fixed effects, $\epsilon$ is the normally-distributed errors, and $f_t$ represents the segmented effect for $t$, with knots at -10\textdegree, 0\textdegree, 5\textdegree, 10\textdegree, 15\textdegree, 20\textdegree, and 25\textdegree C WBGT. We estimate the 95\% confidence interval of all our models using 80-fold bootstrapping.

>>>>

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{Environmental Studies Program, University of Colorado Boulder}
\affil[3]{Department of Geographical Sciences, University of Maryland College Park}
\affil[4]{CSIR-Science and Technology Policy Research Institute, Ghana}
\affil[5]{School of Business, East China University of Science and Technology}
\affil[6]{University of Colorado Boulder}
\affil[7]{University of British Columbia}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

@@

%Theoretical: frameworks and pathways
Researchers have proposed a number pathways by which temperature can affect mental health \cite{Berry2018Apr, Palinkas2020Apr, BerryETAL2010}. Work on biological mechanisms emphasizes the mental health effects from dehydration and other consequences of maintaining stable body temperature in high heat \cite{Lohmus2018Jul, sadiq_impact_2019}. Other studies point to linkages between increased temperatures, overall physical health and increases in injuries which can in turn affect mental health \cite{Berry2007, WHO2007}. Additionally, exposure to higher temperatures can affect productivity and income \cite{kjellstrom_impact_2016, Burke2015Nov}, leading to second-order impacts on mental health outcomes \cite{Katz1997, CohnETAL2004, BouchamaETAL2007}.

%Empirical: Heat and mental health
Drawing on these frameworks, results from an emerging literature are beginning to document linkages between higher temperatures worsened mental health outcomes. Multiple studies find strong evidence that higher temperatures are associated with increases in suicides in the United States \cite{Burke2018Aug, Mullins2019Dec, Dixon2007May} with other studies finding a similar relationship in locations around the world \cite{Qi2014Dec, Page2007Aug, Likhvar2011Jan}. Higher temperatures are also correlated with increased hospitalizations related to general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} and incidents related specific conditions like bipolar disorder and schizophrenia \cite{Lee2007Jan, Sung2013Feb}. Higher temperatures are also linked to increased mortality among people with pre-existing mental health disorders \cite{Hansen2008Oct}.  Finally, recent studies find nighttime temperatures affect sleep quality, with consequences for mental health \cite{Obradovich2017May, Mullins2019Dec}, although these studies have relied on retrospective survey data and have not used data from across the diurnal cycle.

@@

While public health data on mental health outcomes is not available at the neighborhood scale, spontaneous, in situ data from the social media platform Twitter can provide an important, inexpensive and timely proxy for mental health at extremely precise spatial and temporal scales. Many studies use the mood expressed in social media posts as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}. For example, studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder (PTSD) in individuals even before their formal diagnosis \cite{Reece2017Oct}. Similar methods applied to posts on Facebook can predict the onset of depression \cite{Eichstaedt2018Oct}. In London, day-to-day changes in mental health indicators derived from the text of tweets were associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}. Across the USA, more positive tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}, and twitter has provided insight into population-level patterns of happiness over the course of the day \cite{Dodds2011, Dzogang2017}. Finally, the mood expressed in tweets has been strongly associated with local weather conditions \cite{baylis_weather_2018, hannak_tweetin_2012} and been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}. Thus, posts on Twitter provide an proxy for mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures. To explore how neighborhood conditions affect the relationship between heat and mental heath, we draw on a data set of 243 million geolocated tweets from across the continental United States covering the period 2009-2019, each paired with data on local weather and neighborhood conditions.

@@

\begin{minipage}{.65\textwidth}
\centering
\includegraphics[width=\textwidth]{../res/wbgt-income-race.png}
\captionof{figure}{Relationship between WBGT and mood from text in tweets, moderated by the interaction of race and income.}
\label{fig:race-income}
\end{minipage}\hfill
\begin{minipage}{.35\textwidth}
\centering
\begin{tabular}{| l | l |}
\hline
Terms & AIC \\
\hline
Neither & 277,270,561 \\
Income & 277,182,239 \\
Race & 277,143,579 \\
Both & 277,108,676 \\
\hline
\end{tabular}
\captionof{table}{Akaike information criterion (AIC) of models with interaction terms for race, income, both race and income, as well as neither race nor income. The model with the best fit by AIC includes both race and income interaction terms.}
\label{tab:aics}
\end{minipage}

Graphing the combined effects of race and income, we find that race has a larger impact on vulnerability than income (see Fig. \ref{fig:race-income}). A Black neighborhood at any income level will be much more impacted by heat than a poor neighborhood with few black inhabitants. Additionally, neighborhoods that are largely black are also more affected by cold temperatures, although there is no clear effect of income moderating the relationship between cold temperatures and mood. Using AIC as an indicator of how different terms affect model quality (see Tab. \ref{tab:aics}), we find that a model including both terms performs the best, indicating that both race and income matter for the impact of heat on mood. However, including only a race term improves the model more than including only an income term, further supporting the conclusion that race is more important than income for explaining the vulnerability of mood to heat.

@@

\begin{figure}[H]
 \centering
 \includegraphics[width=0.66\linewidth]{../res/comparison_plot.png}
 \caption{Decline in mood as temperature changes from mild (5\textdegree C WBGT) high (25\textdegree C WBGT) across different neighborhood types, compared with the impacts of the change from Saturday to Monday, as well as the change associated with Hurricane Sandy.}
 \label{fig:compare}
\end{figure}

@@

One caveat associated with this data is that we were only able to locate the tweets within the census block where the tweet was sent - we did not to infer where the person sending the tweets typically lived or where they had been, so these results may in fact under-estimate the impacts of higher temperatures on mental health for poor and minority individuals. Moreover, in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are typically found that census block. For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day. Additionally, neighborhoods of young college students are also typically estimated to have low income levels, even through college students are wealthier than the average American. Again, these issues mean we are likely under-estimating the true effect of heat on mental health for poor and Black people. A final issue with using Twitter data is that, while Twitter is used by more than one in five Americans and at similar rates across racial groups, Twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}. Nevertheless, we found a large volume of tweets across all neighborhood types.

@@

Where $y$ is the mood of a tweet, $t$ is the wet bulb globe temperature at the hour of the tweet, $p$ is a binary variable indicating whether it rained at the hour of the tweet, $s$ is the incoming shortwave radiation, or sunshine, in $W/m^2$, at the hour of the tweet, $\Phi$ is the spatio-temporal fixed effects, $\epsilon$ is the normally-distributed errors, and $f_t$ represents the segmented effect for $t$, with knots at -10\textdegree, 0\textdegree, 5\textdegree, 10\textdegree, 15\textdegree, 20\textdegree, and 25\textdegree C WBGT. We estimate the 95\% confidence interval of all our models using 80-fold bootstrapping.
